---
title: "Glioma scRNA project - Loading data per sample"
author: "Youri Hoogstrate"
date: "`r BiocStyle::doc_date()`"
---

```{r, eval=FALSE}
remove.packages('SoupX')
xfun::install_github("https://github.com/yhoogstrate/SoupX/tree/passthrough_arguments")
```


# setup

```{r, message=F, warnings=F}
library(dplyr)

print(getwd()) # 

source('scripts/theme_youri.R')
```

# add sample desrciptors
```{r init df}
data_per_sample <- data.frame()
```

## van Hijfte - GBM sample Y

```{r attach dataset}
data_per_sample <- data_per_sample |> 
  rbind(data.frame(
    "sid" = "van_Hijfte_GBM_Y",
    "sample_name" = "van Hijfte - GBM Sample Y",
    "doi" = "https://dx.doi.org/zenodo.6546712",
    "path_seurat" = 'data/van_Hijfte_GBM_Y/Glioma_Y_and_O/Levi2_Glioma_Y/outs/raw_feature_bc_matrix'
  ))
```

# Add some columns based on existing

```{r}
data_per_sample <- data_per_sample |> 
  dplyr::mutate(path_soupx = gsub("^(.+)/[^.]+$","\\1", path_seurat)) |> 
  dplyr::mutate(n.barcodes = NA) |> 
  dplyr::mutate(n.genes = NA) |> 
  dplyr::mutate(n.reads = NA) |> 
  # cellranger csv
  dplyr::mutate(path_cellranger_summary = paste0(path_soupx, "/metrics_summary.csv"))|> 
  dplyr::mutate(tmp = read.csv(path_cellranger_summary)) |> 
  tidyr::unnest(tmp)
```



# Fig 1: Sample overview

```{r dataset overview}
data_per_sample
```

# obtain generic stats
## SoupX analysis

https://github.com/constantAmateur/SoupX

```{r}
sample <- data_per_sample |> 
  dplyr::slice(1)
```

```{r, message = FALSE, warning=FALSE}
sc <- SoupX::load10X(sample$path_soupx, verbose=T, keepDroplets=TRUE)
SoupX::plotDropletResolution(sc) +
  ggplot2::labs(subtitle = paste0("UMIs per droplot: ", sample$sample_name))
```


## obtain spike-ins

Spike-in virtual soup cells to show where they appear
```{r}
si <- SoupX::spikeInTheSoupChannel(sc,)
saveRDS(si, file=paste0("cache/soup-spike-ins/soup_spike_ins__",sample$sid,"__5000.Rds"))
rm(si)
```

# Select QC passing cells
```{r}
sc <- Seurat::Read10X(data.dir = sample$path_seurat)
si <- readRDS(file=paste0("cache/soup-spike-ins/soup_spike_ins__",sample$sid,"__5000.Rds"))
sc <- cbind(sc, si)
rm(si)
```

create full seurat object, filter later on more
```{r}
sc <- Seurat::CreateSeuratObject(counts = sc,
  min.cells = 3,
  min.features = 150,
  project = sample$sample_name)
sc[['soup.droplets']] <- ifelse(grepl("Soup", Seurat::Cells(sc)),"Soup Spike-in","actual droplet")
```

Add % mito-reads to each droplet

```{r}
mito.features <- grep(pattern = "^MT-", x=rownames(x=sc), value=T)
sc[['percent.mito']] <- Matrix::colSums(x = GetAssayData(object = sc, slot="counts")[mito.features,]) / Matrix::colSums(x = GetAssayData(object = sc, slot = "counts"))
rm(mito.features)
```

```{r}
ribo.features <- grep(pattern = "^RPL", x=rownames(x=sc), value=T)
sc[['percent.ribo']] <- Matrix::colSums(x = GetAssayData(object = sc, slot="counts")[ribo.features,]) / Matrix::colSums(x = GetAssayData(object = sc, slot = "counts"))
rm(ribo.features)
```

as MALAT1 is highly present in soup, see if this is more than anything else
```{r}
sc[['percent.malat1']] <-  GetAssayData(object = sc, slot="counts")[c("MALAT1"),] / Matrix::colSums(x = GetAssayData(object = sc, slot = "counts"))
```

plot qc
```{r}
VlnPlot(object = sc, features = c("nFeature_RNA", "nCount_RNA", "percent.mito", "percent.ribo"), ncol = 4, pt.size = 0.01, group.by = "soup.droplets")
```


```{r}
ggplot(sc@meta.data, aes(y=`nFeature_RNA`, x=orig.ident)) +
  geom_jitter(cex=0.01) +
  geom_hline(yintercept = 1400,col="red") +
  geom_hline(yintercept = 4500,col="red")

ggplot(sc@meta.data, aes(y=`nCount_RNA`, x=orig.ident)) +
  geom_jitter(cex=0.01)  +
  geom_hline(yintercept = 1600,col="red") +
  geom_hline(yintercept = 14000,col="red")
  #scale_y_log10()

sc <- subset(x = sc, subset = 
                     nFeature_RNA > 750 &
                     nCount_RNA > 1400
                     #nFeature_RNA < 4500 & 
                     #nCount_RNA < 14000 &
                     #percent.mito < 0.025
)

sc <- NormalizeData(object = sc, normalization.method = "LogNormalize", scale.factor = 1e4)
sc <- FindVariableFeatures(object = sc, selection.method = "vst", nfeatures = 2000)
sc <- ScaleData(sc)
```

PCA
```{r}
sc <- RunPCA(sc, features = VariableFeatures(object = sc))
VizDimLoadings(sc, dims = 1:2, reduction = "pca")
DimPlot(sc, reduction = "pca",group.by = "soup.droplets")
DimPlot(subset(sc, soup.droplets == "actual droplet"), reduction = "pca") + DimPlot(sc, reduction = "pca",group.by = "soup.droplets")
```


Clustering
```{r}
ElbowPlot(sc, ndims = 45)
sc <- FindNeighbors(sc, dims = 1:34)
sc <- FindClusters(sc, resolution = 0.8, algorithm=1)
sc[['label']] = paste0(sc[['seurat_clusters']]$seurat_clusters ,"_",sc[['soup.droplets']]$soup.droplets)
```


UMAP
```{r}
sc <- RunUMAP(sc, dims = 1:34)
```

VIS umap etc
```{r}
DimPlot(sc, reduction = "umap", label = TRUE, pt.size = .8, group.by = "seurat_clusters")
DimPlot(sc, reduction = "umap", label = TRUE, pt.size = .8, group.by = "label")
DimPlot(sc, reduction = "umap", label = TRUE, pt.size = .8, group.by = "soup.droplets")

sum(sc[['seurat_clusters']]$seurat_clusters == 12 & sc[['soup.droplets']] == "actual droplet")
sum(sc[['seurat_clusters']]$seurat_clusters == 24) # 31 t-cells
sum(sc[['seurat_clusters']]$seurat_clusters == 24  & sc[['lowres']] == T) # 31 t-cells


VlnPlot(sc, features = c('nCount_RNA'))
VlnPlot(sc, features = c('percent.mito'))
VlnPlot(sc, features = c('percent.ribo'))



sc[['lowres']] <- sc[['nFeature_RNA']]$nFeature_RNA < 1400 |
  sc[['nCount_RNA']]$nCount_RNA < 2200 |
  sc[['percent.mito']]$percent.mito >= 0.025
DimPlot(sc, reduction = "umap", label = TRUE, pt.size = .8, group.by = "lowres")
DimPlot(sc, reduction = "umap", label = TRUE, pt.size = .8, group.by = "seurat_clusters")


DimPlot(sc, reduction = "pca", dims=c(1,2), label = TRUE, pt.size = .8, group.by = "label")
DimPlot(sc, reduction = "pca", dims=c(1,3), label = TRUE, pt.size = .8, group.by = "label")
DimPlot(sc, reduction = "pca", dims=c(2,3), label = TRUE, pt.size = .8, group.by = "label")
DimPlot(sc, reduction = "pca", dims=c(1,4), label = TRUE, pt.size = .8, group.by = "label")
DimPlot(sc, reduction = "pca", dims=c(1,5), label = TRUE, pt.size = .8, group.by = "label")


m.11 <- FindMarkers(sc, ident.1 = c('11_Soup Spike-in'), group.by = "label")
m.15 <- FindMarkers(sc, ident.1 = c('15_actual droplet'), group.by = "label")
m.16 <- FindMarkers(sc, ident.1 = c('16_actual droplet'), group.by = "label")
m.23 <- FindMarkers(sc, ident.1 = c('23'), group.by = "seurat_clusters") # hoop low-res

head(m.11, 15)
head(m.15, 15)
head(m.16, 15)


FeaturePlot(sc, c('CSF3R'))
FeaturePlot(sc, c('CD163'))
FeaturePlot(sc, c('PDGFRB'))
FeaturePlot(sc, c('EGFR'))
FeaturePlot(sc, c('RBFOX3'))

FeaturePlot(sc, c('COL1A1'))
FeaturePlot(sc, c('COL1A2'))
FeaturePlot(sc, c('RGS5'))
FeaturePlot(sc, c('CD248'))

FeaturePlot(sc, c('LCK'))
FeaturePlot(sc, c('CD2'))
```


Find soup component
```{r}
DimHeatmap(sc, dims = c(1, 2))
DimHeatmap(sc, dims = c(3, 4))
DimHeatmap(sc, dims = c(5, 6))
DimHeatmap(sc, dims = c(7, 8))
DimHeatmap(sc, dims = c(9, 10)) # PC-9 [juist laag in spike-ins?]?

FeaturePlot(sc, c('JCHAIN','IGHG1'))
```

```{r}
VizDimLoadings(sc, dims = 9, reduction = "pca")
sc[['PC_9']] = FetchData(sc, vars = c("PC_9")) * -1 # was negatieve pc
sc.n <- ScaleData(sc, vars.to.regress = c("PC_9"), features = rownames(sc))
```













```{r}
sc <- SoupX::autoEstCont(sc)
out <- SoupX::adjustCounts(sc)
```

```{r, contribution genes to soup}
plt <- sc |>
  purrr::pluck('soupProfile') |> 
  dplyr::arrange(desc(est)) |> 
  tibble::rownames_to_column('gene_symbol') 

ggplot(plt |> head(n=20), aes(x=reorder(gene_symbol, desc(counts)), y=counts)) +
  geom_point() +
  scale_y_log10() +
  theme_youri_barplot
```

```{r}
SoupX::plotMarkerDistribution(sc)
```


```{r, eval=FALSE}
plotChangeMap(sc, out, "COL1A1")
plotChangeMap(sc, out, "COL1A2")
plotChangeMap(sc, out, "CD248")
plotChangeMap(sc, out, "PDGFRB")
plotChangeMap(sc, out, "CD163")
plotChangeMap(sc, out, "RBFOX3")
SoupX::plotMarkerMa
```


